### 2023-06-18 manuscript/chapters/1_intro.qmd
valentin proofread and comments

Early recordings of the cat's somatosensory visual cortices suggested that the cortical sheet is organized in vertical columns of functionally similar neurons @Mou57.

@@

As models of spatiotemporal neural population dynamics on the cortical sheet, neural field models can be used to explain experimentally observed patterns of cortical activity, such as travelling waves in visual [@SatNau12;@MulRey14], somatosensory [@PetHah03;@FerBol06], motor [@RubRob06;@TakKim15], and hippocampal [@LubSia09;@PatFuj12;@PatSch13] cortices (see @MulCha18 for a review), and are used to model large-scale brain signals, such as electroencephalography (EEG) recordings @Bre17. See also @DipRan18 for an example of neural field modelling of calcium imaging recording of the visual cortex.

@@

The goal of this Master Thesis is to show that the answer to the second question is highly nontrivial because even the dimensionality of the embedding space, for some given spatiotemporal dynamics, is not clearly defined. This means that it is possible to find two embeddings (and connectivity kernels) of different dimensions that give rise to identical dynamics; *contradicting the intuition that the embedding dimension should at least be as high as the dimensionality of the dynamics* (the number of variables needed to fully describe the dynamical system).

In particular, we show that for any two-dimensional neural field equation (that is, with two spatial dimensions), there is a one-dimensional neural field equation on the interval, where the embedding space is simply the interval $[0,1]$, from which the solution to the two-dimensional equation can be fully reconstructed. The mapping from the two-dimensional equation to the one-dimensional equation is done via well-known measurable bijections between the square $[0,1]^2$ and the interval $[0,1]$. Importantly, these bijections are not (and can't be) diffeomorphisms. Also, we show that although the connectivity kernel of the one-dimensional equation seems to have a fractal structure, it is sufficiently regular for its solution to be numerically approximated using standard mesh-based simulations.

@@

Let us now give a short introduction to the mathematical formulation of neural fields. Neural field equations emerge from the structure of networks of rate neurons when the number of neurons $N$ becomes large. Positions $\vec z_i \in \Rp$ are attributed to each neuron, which defines the *embedding space*. Each neuron has a potential $h_i(t)$ that evolves in time, and the weight with which the neurons interact with each other is given by the components of the connectivity matrix $J_{ij} \in \R^{N \times N}$. Additionally, the activation function $\phi(h)$ describes the firing rate of a given neuron with potential $h$. Then, the product $J_{ij} \phi(h_j)$ gives the *recurrent current* of neuron $j$ on neuron $i$.

@@

If the embedding is well-chosen,[^1] the connectivity matrix $J_{ij}$ approaches a connectivity kernel $w(\vec y, \vec z)$ as $N \rightarrow \infty$, and neurons close in the embedding space tend to have similar potentials, such that we can write the continuous *neural field* $h(\vec z, t)$ of neuron potentials. Before we write down the neural field equation which describes how the field evolves in time, we need to introduce the neuron density $\rho(\vec z)$, which measures the number of neurons per unit volume of the embedding space. The neural field equation is then written as such:

@@

![(TODO add phi, add z instead of xi) The connectivity kernel defines the influence between neural densities in the embedding space](figures/connectivity_kernel.png)

@@

This work was inspired by recent mathematical development in mean-field theories involving the theory of graphons @JabPoy21. Their main result is that under some boundedness assumptions of the connectivity between the components of a system, the equation of evolution incorporates only an integral over $[0,1]$. This is indeed the case in the toy model described next chapter, where weights scale as $J_{ij} \sim \frac 1N$, which motivates the search for a formulation of a one-dimensional neural embedding.

The main idea is to define a measurable bijective mapping $S : \Rp \mapsto [0,1]$ and a measure $\lambda : [0,1]^2 \mapsto \mathbb{R}^+$, such that we can write a one-dimensional neural field	$\tilde h(\alpha, t)$ with a connectivity kernel loosely defined as $\tilde w = w \circ S^{-1}$.

@@

This change of variable in the kernel seems counter-intuitive: how is it possible to map a higher dimension to a lower dimension without losing information? And even if mathematically such a change of variables is possible, is it even feasible to simulate it numerically?

@@

1. A neuron-based simulation of $N$ neurons. Each neuron samples the probability distribution $\rho(z_1, \cdots, z_p)$ independantly. This case then just reduces to simulating the network of rate neurons. In other words, we are simulating the full network of rate neurons, and the simulation converges to the true mathematical neural field equation as $N$ becomes large.

@@

With this in mind, we can try to understand why simulating a neural field in $[0,1]$ might be difficult. Suppose we take bins in the $\Rp$ space, and map each bin to a random position in $[0,1]$. Intuitively, this won't work (in the sense that the dynamics in $[0,1]$ will not be the same as the dynamics in $\Rp$), because the hypothesis of continuity is not longer true: the mapping is random, so the nearby populations in $[0,1]$ will no longer have similar potentials, and the error of the mean-field approximation will no longer vanish as we take increasingly many bins.

>>>>

Early recordings of cat's somatosensory @Mou57 and visual @HubWie62 cortices suggested that the cortical sheet is organized in vertical columns of functionally similar neurons.

@@

As models of spatiotemporal neural population dynamics on the cortical sheet, neural field models can be used to explain experimentally observed patterns of cortical activity, such as travelling waves in visual [@SatNau12;@MulRey14], somatosensory [@PetHah03;@FerBol06], motor [@RubRob06;@TakKim15], and hippocampal [@LubSia09;@PatFuj12;@PatSch13] cortices (see @MulCha18 for a review), and are used to model large-scale brain signals, such as electroencephalography (EEG) recordings @Bre17 (see also @DipRan18 for an example of neural field modelling of calcium imaging recording of the visual cortex).

@@

The goal of this Master Thesis is to show that the answer to the second question is highly nontrivial because even the dimensionality of the embedding space, for some given spatiotemporal dynamics, is not clearly defined. This means that it is possible to find two embeddings (and connectivity kernels) of different dimensions that give rise to identical dynamics.
<!-- ; *contradicting the intuition that the embedding dimension should at least be as high as the dimensionality of the dynamics* (the number of variables needed to fully describe the dynamical system). -->

In particular, we show that for any two-dimensional neural field equation (that is, with two spatial dimensions), there is a one-dimensional neural field equation, where the embedding space is simply the interval $[0,1]$, from which the solution to the two-dimensional equation can be fully reconstructed. The mapping from the two-dimensional equation to the one-dimensional equation is done via well-known measurable bijections between the square $[0,1]^2$ and the interval $[0,1]$. Importantly, these bijections are not (and can't be) diffeomorphisms. Also, we show that although the connectivity kernel of the one-dimensional equation seems to have a fractal structure, it is sufficiently regular for its solution to be numerically approximated using standard grid-based simulations.

@@

Let us now give a short introduction to the mathematical theory of neural fields. Neural field equations emerge from the spatial structure of networks of rate neurons (or Poisson spiking neurons [@Ger95;@CheDua19]) when the number of neurons $N$ becomes large. Positions $\vec z_i \in \Rp$ are attributed to each neuron, where $\Rp$ is the *embedding space*. Each neuron has a potential $h_i(t)$ that evolves in time, and the weight with which the neurons interact with each other is given by the entries of the connectivity matrix $J_{ij} \in \R^{N \times N}$. Additionally, the activation function $\phi(h)$ describes the firing rate of a given neuron with potential $h$. Then, the product $J_{ij} \phi(h_j)$ gives the *recurrent current* of neuron $j$ on neuron $i$.

@@

If the embedding is well-chosen,[^1] the connectivity matrix $J_{ij}$ approaches a connectivity kernel $w(\vec y, \vec z)$ as $N \rightarrow \infty$, and neurons close in the embedding space tend to have similar potentials, such that we can write the continuous *neural field* $h(\vec z, t)$ of neuron potentials. Before we write down the neural field equation which describes how the field evolves in time, we need to introduce the neuron density $\rho(\vec z)$, which measures the number of neurons per unit volume of the embedding space. The neural field equation writes:

@@

![(TODO add phi, add z instead of xi) The connectivity kernel defines the influence between localized neural populations in the embedding space](figures/connectivity_kernel.png)

@@

This work was inspired by recent mathematical development in mean-field theories involving the theory of graphons @JabPoy21.
Informally, these recent results suggest that, even in the absence of any explicit spatial structure, as long as weights scale as $1/N$, the dynamics of large networks can be described, as $N \to \infty$, by a neural field equation where the embedding is simply the interval $[0,1]$, which motivates the search for a formulation of a one-dimensional neural embedding.
<!-- Their main result is that under some boundedness assumptions of the connectivity between the components of a system, the equation of evolution incorporates only an integral over $[0,1]$. This is indeed the case in the toy model described next chapter, where weights scale as $J_{ij} \sim \frac 1N$, which motivates the search for a formulation of a one-dimensional neural embedding. -->

The main idea is to define a measurable bijective mapping $S : \Rp \mapsto [0,1]$ and a measure $\lambda : \Rp \mapsto \mathbb{R}^+$, such that we can write a one-dimensional neural field	$\tilde h(\alpha, t)$ with a connectivity kernel loosely defined as $\tilde w = w \circ S^{-1}$.

@@

This change of variable in the kernel seems counter-intuitive: how is it possible to map dynamics in the embedding space $\Rp$ to dynamics in the lower dimensional embedding space $[0,1]$ without losing information? And even if mathematically such a change of variables is possible, is it even feasible to numerically simulate the dynamics in the lower dimensional embedding space?

@@

1. A neuron-based simulation of $N$ neurons. The location of each neuron is an independent sample from the distribution $\rho(z_1, \cdots, z_p)$. This case then just reduces to simulating the network of rate neurons. In other words, we are simulating the full network of rate neurons, and the simulation converges to the true mathematical neural field equation as $N$ becomes large.

@@

With this in mind, we can try to understand why simulating a neural field in $[0,1]$ might be difficult. Suppose we take bins in the $\Rp$ space, and map each bin to a random position in $[0,1]$. Intuitively, this won't work (in the sense that the dynamics in $[0,1]$ will not be the same as the dynamics in $\Rp$), because the hypothesis of continuity is not longer true: the mapping is random, so the nearby neurons in $[0,1]$ will no longer have similar potentials, and the error of the mean-field approximation will no longer vanish as we take increasingly many (and smaller) bins.
<!-- NOTE : was "populations" before, this is explained later in the thesis. but OK, for the introduction, maybe we should use neurons
the mapping is random, so the nearby populations in $[0,1]$ -->

### 2023-06-18 manuscript/chapters/2_neural_fields.qmd
dkfjwen fhjk !!!!!
valentin proofread and comments

The toy model we introduce is an instance of a low-rank rate neural network. Such models hold their name from the form of their connectivity matrix, which has a rank of (at most) $p$. We write the connectivity matrix as

@@

The geometric view of these networks is that the recurrent currents lie in a $p$-dimensional subspace of $\R^N$ spanned by the vectors $\{\vec{F_1}, \cdots, \vec{F_p}\}$ (where $\vec{F_\mu}=(F_{\mu 1}, \cdots F_{\mu N})$), which therefore form a "natural" embedding.

Recalling the notation from the introduction, $h_i(t)$ are the neuron potentials (with initial condition $h_i(0)$), $\phi : \R \mapsto \R^+$ is the monotonic increasing activation function, and we write the evolution equation for this network of neurons:

@@

The exponential decay term describes how, in isolation, a neuron's potential tends to zero, biologically corresponding to a depolarized membrane. The recurrent current term $I^\text{rec}_i(t)$ describes the "currents" received by neuron $i$ from all the other neurons in the network.

@@

One last remark is that in our model, we omit to add any source of external currents: from the initial condition, the network is left to evolve in isolation.

@@

Let us now specify the low-rank model introduced in the previous section, and use the model introduced in @schmutz2023convergence. This paper defines the low-rank vectors $F_{\mu i}$ such that each component independently samples a standard Gaussian (zero mean, unit variance). In other words, every neuron $i$ samples a vector $\vec{F_i} = (F_{1i}, \cdots, F_{pi})$ from the $p$-dimensional gaussian ball. We write

@@

$$
G_{\mu i} = \tilde\phi(z_{\mu i}) \stackrel{\text{def}}{=} \frac{\phi(z_{\mu i}) - \avg{z_{\mu i}}}{\mathrm{Var}[\phi(z_{\mu i})]}.
$$

This choice be elucidated in a few paragraphs. Finally, the activation function remains to be defined. For simplicity, we take it to be the logistic function, although the following results are not sensitive to this choice.

$$
\phi(h) = \frac{1}{1 + \mathrm{e}^h}
$$

@@

$$
\dot h_i(t) = -h_i(t) + \frac 1N \sum_{\mu=1}^p \sum_{j=1}^{N} z_{\mu i} \tilde \phi(z_{\mu j}) \phi(h_j(t)).
$$ {#eq-toymodel-evolution}

@@

$$
\dot h_i(t)\ \text{evaluated at}\ h_i^\star = -h^\star_i(t) + \frac 1N \sum_{\mu=1}^p \sum_{j=1}^{N} z_{\mu i} \tilde \phi(z_{\mu j}) \phi(h^\star_j(t)) = 0
$$

@@

In the introduction, we introduced the convergence of a network of neurons to a smooth neural field when the embedding space is well-chosen. For our toy model, it seems natural to try to embed the neurons according to their positions on the $p$-dimensional Gaussian ball.

@@

The results of the simulation show that the random ordering of the neurons does not yield a smooth embedding, on contrary of the smooth surface that appears when we order the neurons by their position $z_{1i}$. The simulation serves as a demonstration that the neurons converge to the stable fixed point $h^\star_i = z_{1i}$, as well as confirming that the zero fixed point is unstable. Indeed, despite the zero initialization, the simulation did not stay there because small numerical fluctuations were amplified by the instability.

@@

### Sampling method (Monte Carlo integration)

Due to the formulation of the neural field weighed by a probability distribution $\rho(y_1, \cdots, y_p)$, a natural way to estimate the integral is by Monte Carlo integration.
The essence of this method is the use of the Central Limit Theorem, where we take $N$ independent samples $\vec{y_i} = (y_{1i}, \cdots, y_{pi})$ from the distribution $\rho$, and use them to estimate the integrand.

@@

The attentive reader might have noticed by now that the Monte-Carlo method just reduces to simulating the network of neurons @eq-toymodel-evolution. Effectively, the network of neurons converges to the neural field with rate $\bO(1/\sqrt{N})$.

One small nuance we should add, is that strictly speaking, Monte-Carlo integration should resample the integrand at every timestep, but the network of neurons samples the neurons once, and propagates these points through time.

@@

Overlap variables were introduced by @Gerstner_1992 in the context of the study of the Hopfield network, which can be seen as a low-rank network with discrete variables instead of our continuous $\vec{z}$. The overlap variables measure the correlation between the state of the network (the value of the potential $h(\vec z, t)$ at each point in space) and the pattern fixed point.

@@

We cite @veltz2009localglobal, section 4.3 for the reduction of general low-rank networks to $p$-dimensional closed systems, and give the equations applied to our toy model. Defining the projections $\kappa_\mu(t)$ of the neural field onto the patterns

@@

In @sec-networks-of-neurons-to-neural-fields we introduced a simple toy model a neural field with a natural embedding on the $p$-dimensional Gaussian ball. Analytical derivations accompanied by numerical simulations showed the dynamics of this model can be summarized by the convergence to pattern fixed points. In this section, we modify the toy model minimally, such that we can observe more interesting behaviour in the form of cycling.

@@

1. We define a time delay $\delta$ with which a neuron will "wait" before reacting to a change in its potential.
2. We define "rolling" as the response of a neuron to shift towards the "next" pattern given its state matching a current pattern. This is done by adding a shift $\mu+1$ in the connectivity kernel, with the convention $p+1=1$.

@@

![(TODO animation in 3D, rename non-cycling to original) Intuition for the cycling behavior by considering the recurrent currents](figures/cycling_recurrent.png){#fig-cycling-irec}

@@

With these considerations, we simulate a $p=2$ cycling neural field with $\delta=6$ and initial condition $h(z_1, z_2, t < 0) = z_1$ in @fig-cycling-anim. The latent trajectory is estimated from the simulation results, and animated simultaneously with the neural field. We clearly see the oscillations between the patterns $\mu=1$ and $\mu=2$.

@@

This system of equations can be solved by steps of $\delta$, iteratively plugging in the previous solution to solve each cycle.

@@

Defining the change of variables $v_\mu=\CDF(z_\mu), u_\mu=\CDF(y_\mu)$, we can define the uniform neural field and uniform connectivity kernel as

>>>>

The toy model we introduce is an instance of a low-rank neural network. Such models hold their name from the form of their connectivity matrix, which has a rank of (at most) $p$. We write the connectivity matrix as

@@

The geometric view of these networks is that the recurrent currents lie in a $p$-dimensional subspace of $\R^N$ spanned by the vectors $\{\vec{F_1}, \cdots, \vec{F_p}\}$ (where $\vec{F_\mu}=(F_{\mu 1}, \cdots F_{\mu N})$), which therefore define a "natural" linear embedding (in the sense of @JazOst21) of the neural population activity.

Recalling the notation from the introduction, the $h_i(t)$ are neuron potentials (with initial condition $h_i(0)$), $\phi : \R \mapsto \R^+$ is the monotonic increasing activation function, and we write the evolution equation for this network of neurons:

@@

The exponential decay term describes how, in isolation, a neuron's potential tends to zero, biologically corresponding to the decay of the membrane potential to its value at rest. The recurrent current term $I^\text{rec}_i(t)$ describes the "currents" received by neuron $i$ from all the other neurons in the network.

@@

One last remark is that in our model, there is no external current; we only study the autonomous dynamics of the system.

@@

Let us now specify the low-rank model introduced in the previous section, and use the model introduced in @schmutz2023convergence, which can be seen as a simple example of a "Gaussian mixture low-rank network" @Beiran2021. This paper defines the low-rank vectors $F_{\mu i}$ such that each component independently samples a standard Gaussian (zero mean, unit variance). In other words, every neuron $i$ samples a vector $\vec{F_i} = (F_{1i}, \cdots, F_{pi})$ from the $p$-dimensional gaussian. We write

@@

$$
G_{\mu i} = \tilde\phi(z_{\mu i}) \stackrel{\text{def}}{=} \frac{\phi(z_{\mu i}) - \avg{\phi(z_{\mu i})}}{\mathrm{Var}[\phi(z_{\mu i})]}.
$$

The motivation for this choice will be presented in a few paragraphs. Finally, the activation function remains to be defined. For simplicity, we take it to be the logistic function, although the following results are not sensitive to this choice,

$$
\phi(h) = \frac{1}{1 + \mathrm{e}^h}.
$$

@@

$$
\dot h_i(t) = -h_i(t) + \frac 1N \sum_{\mu=1}^p \sum_{j=1}^{N} z_{\mu i} \tilde \phi(z_{\mu j}) \phi(h_j(t)) =: \mathrm{RHS}(h_1, \cdots, h_N).
$$ {#eq-toymodel-evolution}

@@

<!--
$$
\dot h_i(t)\ \text{evaluated at}\ h_i^\star = -h^\star_i(t) + \frac 1N \sum_{\mu=1}^p \sum_{j=1}^{N} z_{\mu i} \tilde \phi(z_{\mu j}) \phi(h^\star_j(t)) = 0
$$ 
-->

$$
\mathrm{RHS}(h^\star_1, \cdots, h^\star_N) = -h^\star_i + \frac 1N \sum_{\mu=1}^p \sum_{j=1}^{N} z_{\mu i} \tilde \phi(z_{\mu j}) \phi(h^\star_j) = 0
$$

@@

In the introduction, we introduced the convergence of a network of neurons to a smooth neural field when the embedding space is well-chosen. For our toy model, it seems natural to try to embed the neurons according to their positions on the $p$-dimensional Gaussian.

@@

The results of the simulation show that the random ordering of the neurons does not yield a smooth embedding, in contrast to the smooth surface that appears when we order the neurons by their position $z_{1i}$. The simulation serves as a demonstration that the neurons converge to the stable fixed point $h^\star_i = z_{1i}$, as well as confirming that the zero fixed point is unstable. Indeed, despite the zero initialization, the simulation did not stay there because small numerical fluctuations were amplified by the instability.

@@

### Sampling method (Monte Carlo integration) {#sec-mc}

Due to the formulation of the neural field weighed by a probability distribution $\rho(y_1, \cdots, y_p)$, a natural way to estimate the integral is by Monte Carlo integration.
The essence of this method is to use the Central Limit Theorem, where we take $N$ independent samples $\vec{y_i} = (y_{1i}, \cdots, y_{pi})$ from the distribution $\rho$, and use them to estimate the integrand.

@@

<!-- The attentive reader might have noticed by now that the Monte-Carlo method just reduces to simulating the network of neurons @eq-toymodel-evolution. Effectively, the network of neurons converges to the neural field with rate $\bO(1/\sqrt{N})$.

One small nuance we should add, is that strictly speaking, Monte-Carlo integration should resample the integrand at every timestep, but the network of neurons samples the neurons once, and propagates these points through time. -->

@@

Overlap variables in continuous time variables were introduced by @Gerstner_1992 to study spiking networks models of associative memory (Hopfield-type networks), which can be seen as a low-rank network with discrete spatial structure instead of our continuous $\vec{z}$. The overlap variables measure the correlation between the state of the network (the value of the potential $h(\vec z, t)$ at each point in space) and the pattern fixed point.

@@

We refer to @veltz2009localglobal, section 4.3 for the reduction of general low-rank networks to $p$-dimensional closed systems, and give the equations applied to our toy model. Defining the projections $\kappa_\mu(t)$ of the neural field onto the patterns

@@

In @sec-networks-of-neurons-to-neural-fields we introduced a simple toy model of a neural field with a natural embedding on the $p$-dimensional Gaussian. Analytical derivations accompanied by numerical simulations showed the dynamics of this model can be summarized by the convergence to pattern fixed points. In this section, we modify the toy model minimally, such that we can observe more interesting behaviour in the form of cycling.

@@

1. We define a time delay $\delta$ with which a neuron will "wait" before reacting to a change in its potential.
2. We define "rolling" as the response of a neuron to shift towards the "next" pattern given its state matching a current pattern. This is done by adding a shift $\mu+1$ in the connectivity kernel, with the convention $p+1=1$ (see @Ami89, Chapter 5).

@@

![(TODO animation in 3D, rename non-cycling to original, number from one) Intuition for the cycling behavior by considering the recurrent currents](figures/cycling_recurrent.png){#fig-cycling-irec}

@@

With these considerations, we simulate a $p=2$ cycling neural field with $\delta=6$ (since the membrane time constant has been fixed to one, this means that $\delta = 6\tau = 6$) and initial condition $h(z_1, z_2, t < 0) = z_1$ in @fig-cycling-anim. The latent trajectory is estimated from the simulation results, and animated simultaneously with the neural field. We clearly see the oscillations between the patterns $\mu=1$ and $\mu=2$.

@@

<!-- This system of equations can be solved by steps of $\delta$, iteratively plugging in the previous solution to solve each cycle. -->

@@

Defining the change of variables $v_\mu=\CDF(z_\mu), u_\mu=\CDF(y_\mu)$, we can define the neural field $h_U$ and the connectivity kernel $w_U$ for the uniform distribution as
